Refinement of Parallel Algorithms down to LLVM

### 2023-04-11 papers/JAR_SI_ITP2022/main.tex
upd paper

On top of the back-end, a program logic is used to prove programs correct.
It uses separation logic, and provides automation like a verification condition generator (VCG).
In Section~\ref{sec:seplogic}, we describe our formalization of concurrent separation logic~\cite{OH04}, and our VCG.

@@

Using our approach, complex algorithms and data structures can be developed and
refined to optimized efficient code. The stepwise refinement ensures a separation of concerns
between high-level algorithmic ideas and low-level optimizations. We have used this approach to
verify a wide range of practically efficient algorithms~\cite{ELNN13,BrLa18,WiLa18,La17_CADE,La17_SAT,FBL18,La14,LaSe16,LaSe19,La20}.
In Section~\ref{sec:parsort}, we use our techniques to verify a parallel sorting algorithm,
with competitive performance wrt.\ unverified state-of-the-art algorithms.

@@

\subsection{Notation}
We mainly use Isabelle/HOL notation, with some shortcuts and adaptations for presentation in a paper.
In this section, we give examples of the more unusual notations: for implication, we always write \is{==>},
e.g., \is{\<forall>x. x<4 ==> x<=3} (in Isabelle, one must use \is{-->} here).

Type variables are \is{'a, 'b, ...} and a function types are written curried as \is{'a => 'b => 'c}.
Types can be annotated to any symbol or term, e.g., \is{(x::nat) < 4 ==> x<=3}.
Function application is written as \is{f x}, and function update is \is{f(x:=y)}.
For definitions, we use \is{==}, e.g., \is{f x == x+1}.

Algebraic datatypes are written as \is{'a option == Some (the: 'a) | None}. This also defines the selector function \is{the :: 'a option => 'a}.
Values of the tuple type \is{'a_1 \x ... \x 'a_n} are written as \is{(a_1,...,a_n)}. Lists have type \is{'a list},
and we write \is{[]} for the empty list, \is{x#xs} for the list with head \is{x} and tail \is{xs}, and \is{xs_1@xs_2} for list concatenation.
We also use the list notation \is{[1,2,3]}. Then length of list \is{xs} is \is{|xs|}.

The empty set is \is${} :: 'a set$. The set \is${l..<h}$ contains all elements between $l$ inclusive and $h$ exclusive, and the set \is${l..}$ contains all elements greater than or equal to $l$.
Disjoint union is written as \is{$\dot\cup$}. We only write \is{$\dot\cup$} if it is clear from the context where the disjointness constraint belongs to.
Otherwise, we will explicitly write, e.g. \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of the finite set \is{s} is \is{|s|}.

Lambda abstraction is written as \is{\<lambda>x. x+1}. When clear from the context, we omit the \is{\<lambda>}, e.g. \is{spec x. x>42}.
We also use Haskell-like sections for infix operators, e.g. \is{(+1)} for \is{\<lambda>x. x+1}.
Also, we can use underscores to indicate the parameter positions, e.g., \is${_..<_}$ for \is$\<lambda>l h. {l..<h}$.

@@

  Now assume that we have a state (memory) type \is{'\mu}, and an access report type \is{'\rho}, which forms a monoid (\is{0,+}).
  With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
  \begin{lstlisting}
    'x M == '\mu => ('x \x '\rho \x '\mu) neM
    return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
    bind$_M$ m f \mu == (x_1,r_1,\mu) <- m \mu; (x_2,r_2,\mu) <- f x_1 \mu; return$_{ne}$ (x_2,r_1+r_2,\mu)
  \end{lstlisting}
  Here, \is{return} does not change the state, and reports no accesses (\is$0$),
  and \is{bind} sequentially composes the executions, threading through the state $\mu$ and adding up the access reports \is$r_1$ and \is$r_2$.

  Typically, the access report will contain read and written addresses, such that data races can be detected.
  Moreover, if parallel executions can allocate memory, we must detect those
  executions where the memory manager allocated the same block in both parallel strands.
  As we assume a thread safe memory manager, those \emph{infeasible} executions can safely be ignored.
  Let \is{feasible :: '\rho => '\rho => bool} and \is{norace :: '\rho => '\rho => bool} be symmetric predicates,
  and let \is{combine :: ('\rho \x '\mu) => ('\rho \x '\mu) => ('\rho \x '\mu) } be a commutative operator to
  combine two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
  \begin{lstlisting}
    (m_1 || m_2) \mu ==
      (x_1,r_1,\mu_1) <- m_1 \mu; (x_2,r_2,\mu_2) <- m_2 \mu;  --"\rightcomment{execute both strands}"
      assume feasible \rho_1 \rho_2;  --"\rightcomment{ignore infeasible combinations}"
      assert norace \rho_1 \rho_2; --"\rightcomment{fail on data race}"
      return$_{ne}$ ((x_1,x_2), combine (\rho_1,\mu_1) (\rho_2,\mu_2))  --"\rightcomment{combine results}"

    assume P == if P then return () else spec (%_. False)
    assert P == if P then return () else fail
  \end{lstlisting}
  Here, we use \is{assume} to ignore infeasible executions, and \is{assert} to fail on data races.
  Note that, if one parallel strand fails, and the other parallel strand has no possible
  results (\is{spec (\<lambda>_. False)}), the behaviour of the parallel composition is not clear.
  For this reason, we fix an invariant \is{invar$_M$ :: ('\mu => ('x \x '\rho \x '\mu) neM) => bool},
  which implies that every non-failing execution has at least one possible result.
  We define the actual type \is{M} as the subtype satisfying \is{invar$_M$}.
  Thus, we have to prove that every combinator and instruction of our semantics
  preserves the invariant, which is an important sanity check.
  As additional sanity check, we prove symmetry of parallel composition:
  \begin{lstlisting}
    m_1 || m_2 = mswap (m_2 || m_1)     $\text{where}$    mswap m == (x_1,x_2)<-m; return (x_2,x_1)
  \end{lstlisting}
%   where \is|mswap m = (x_1,x_2)<-m; return (x_2,x_1)| swaps the result tuple.

@@

    We then show that \is{malloc} satisfies the invariant of M: we correctly report the allocated block,
    and we can select any fresh block. As our memory model guarantees an infinite supply of fresh blocks,
    any finite set of blocks can be avoided:
    \begin{lstlisting}
      invarM_malloc: invarM (malloc vs)
    \end{lstlisting}

@@

    Indexing and updating of arrays is implemented by:
    \begin{lstlisting}
      aget p i == ll_ofs_ptr p i; ll_load p
      aset p i x == ll_ofs_ptr p i; ll_store x p; return p
    \end{lstlisting}
    The abstract and concrete programs are linked by the following refinement theorems:
    \begin{lstlisting}
      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ (aget xs$_\dagger$ i$_\dagger$)$\hspace*{-1pt}$ (arr$_A$ xs$_\dagger$ xs$\hspace*{1pt}$*$\hspace*{1pt}$idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ id$_A$$\hspace*{-1pt}$ $\top$$\hspace*{-1pt}$ (lget xs i)
      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i) (aset xs$_\dagger$ i$_\dagger$ x) (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r=xs$_\dagger$) (lset xs i x)
    \end{lstlisting}
    That is, if the list \is$xs$ is refined by array \is{xs$_\dagger$}, and the natural number \is$i$ is refined
    by the fixed-width\footnote{We use Isabelle's word library here, which encodes the actual width as a type variable,
    such that our functions work with any bit width. For code generation, we will fix the width to 64 bit.} word \is{i$_\dagger$} (\is{idx$_A$ i$_\dagger$ i}),
    the \is{aget} operation will return the same result as the \is{lget} operation (\is{id$_A$}).
    The resulting memory will still contain the original array. Note that there is no explicit precondition ($\top$) that the array access
    is in bounds, as this follows already from the assertion in the abstract \is{lget} operation.
    The \is{aset} operation will return a pointer to an array that refines the updated list returned by \is{lset}.
    As the array is updated in place, the original refinement of the array is no longer valid.
    Moreover, the returned pointer \is{r} will be the same as the argument pointer \is{xs$_\dagger$}.
    This information is important for refining to parallel programs on disjoint parts of an
    array (cf.\ Section~\ref{sec:array_split}).

@@

  Next, we implement a set as the union of a list of non-empty, pairwise disjoint, and finite sets.
  While that seems to make little sense at first glance, we will later implement the sets in the list by intervals,
  and the list itself by dynamic arrays, to obtain an imperative interval list data structure.
  We define operations for constructing an empty set, emptiness test,
  disjoint union with a single set, cardinality, and a more specialized operation \is{split}, which splits off
  a non-empty set from the list:
  \begin{lstlisting}
    ivl$_R$ :: nat set list => nat set => bool
    ivl$_R$ ls s == s = \<Union> set s & pw_disjoint ls & \<forall>x:set ls. finite x & x ~= {}

@@

  We then refine the lists of sets to array lists (dynamic arrays) of intervals: \is{al$_A$ iv$_A$}.
  Here, \is{al$_A$} is the refinement assertion from lists to the array list data structure from the IRF collections library.
  As argument, it takes a refinement assertion for the list elements.
  Again, Sepref automatically generates imperative implementations of the \is{ivl} functions and proves the corresponding
  refinement lemmas. Combining them with the refinements to sets yields the desired imperative interval list data structure,
  with the refinement relation \is{ivl$_A$ == al$_A$ iv$_A$ O ivl$_R$}.
  For example, for joining a single interval to the list, and for splitting off an interval, we get:
  \begin{lstlisting}
    ivl_dj_un$_\dagger$, ($\dot\cup$) : iv$_A$ -> ivl$_A$^d -> ivl$_A$
    ivl_split$_\dagger$, split : ivl$_A$^d -> iv$_A$ \x ivl$_A$
  \end{lstlisting}
  Note that we update the underlying dynamic array destructively, hence the $\cdot^d$ annotation to the argument refinements.

@@

  Analogously to \is{with_split}, we define a combinator \is{with_idxs s xs f}, that splits
  the list \is{xs} into the lists with the indexes \is{s}, and without the indexes \is{s}, executes \is{f} on these lists,
  and joins the resulting lists:
  \begin{lstlisting}
  with_idxs s xs m ==
    assert (\<forall>i\<in>s. i<|xs| & xs!i ~= None);
    let (xs_1, xs_2) = (sl_split s xs, sl_split (-s) xs);
    (xs$_1'$,xs$_2'$) <- f xs_1 xs_2;
    assert (sl_struct xs$_1'$ = sl_struct xs_1 & sl_struct xs$_2'$ = sl_struct xs_2);
    return (sl_join xs$_1'$ xs$_2'$)
  \end{lstlisting}

@@

  Figure~\ref{fig:psort} show our abstract parallel sorting algorithm \is{psort}.
  This algorithm is derived from the well-known quicksort and introsort algorithms~\cite{Muss97}:
  like quicksort, it partitions the list (line~\ref{l:psort:partition}), and then recursively sorts the partitions in parallel (l.~\ref{l:psort:rec_par}).
  Like introsort, when the recursion gets too deep, or the list too short, we fall back to
  some (not yet specified) sequential sorting algorithm (l.~\ref{l:psort:fallback}). Similarly, when the partitioning is very unbalanced (l.~\ref{l:psort:bad}),
  we sort the partitions sequentially (l.~\ref{l:psort:rec_seq}). These optimizations aim at not spawning threads for small sorting tasks, where
  the overhead of thread creation outweighs the advantages of parallel execution.
  A more technical aspect is the extra parameter \is{n} that we introduced for the list length.
  Thus, we can refine the list to just a pointer to an array, and still access its length\footnote{Alternatively, we could refine a list to a pair of array pointer and length.}.

@@

  Our sorting algorithms are developed and verified in the context of this locale (to avoid confusion, our presentation has, up to now, just used \is{<}, \is{<=}, and \is{sorted} instead of \is{lt p}, \is{le p}, and \is{sorted_wrt (le p)}).
  To get an actual sorting algorithm, we instantiate the locale with an abstract and concrete compare function, proving
  that the abstract function is a weak ordering, and that the concrete function refines the abstract one.
  For our example of sorting indexes into an array, where the array elements themselves are compared by a function \is{lt}, we get:
  \begin{lstlisting}
    interpretation idx: pcmp lt_idx lt_idx$_\dagger$ arr$_A$ idx$_A$   \<proof>

    lt_idx xs i j == lt (xs!i) (xs!j)
    lt_idx$_\dagger$ xs$_\dagger$ i$_\dagger$ j$_\dagger$ == x$_\dagger$<-aget xs$_\dagger$ i$_\dagger$; y$_\dagger$<-aget xs$_\dagger$ j$_\dagger$; lt$_\dagger$ x$_\dagger$ y$_\dagger$
  \end{lstlisting}
  This yields sorting algorithms for sorting indexes, taking an extra parameter for the array to index into.
  For our sampling application, we use \is{idx.introsort xs}.

@@

  \begin{tikzpicture}
    \begin{loglogaxis}[
      xlabel near ticks,
      legend style = {
        at = {(.97,.4)},
        cells={anchor=west},
        font=\footnotesize
      },
      ymajorgrids,
      xmin=10^4,xmax=10^8,
      title={\large Server (u64)},
      title style={at={(0.25,.75)}},
      width=.54\textwidth,
      height=.3\textheight
    ]
      \input{sadata_uint64_server}
    \end{loglogaxis}
  \end{tikzpicture}\hfill
  \begin{tikzpicture}
    \begin{loglogaxis}[
      xlabel near ticks,
      legend style = {
        at = {(.97,.4)},
        cells={anchor=west},
        font=\footnotesize
      },
      ymajorgrids,
      xmin=10^3,xmax=10^7,
      title={\large Server (str)},
      title style={at={(0.25,.75)}},
      width=.54\textwidth,
      height=.3\textheight
    ]
      \input{sadata_llstring_server}
      \legend{}
    \end{loglogaxis}
  \end{tikzpicture}
  \caption{Performance for sorting small arrays of randomly distributed integers and strings.
    The $y$-axis shows the runtime in milliseconds, and the $x$-axis the array size. Note that both axis are logarithmic.
  }\label{fig:smallarrays}

@@

    While the verification of our sorting algorithms uses a top-down approach, we actually started with
    implementing, benchmarking, and fine-tuning the algorithms in C++. This gave us a quick option to
    find an algorithm that is efficient, and, at the same time not too complex for verification,
    without having to verify each intermediate step towards this algorithm.
    Only then we used our top-down approach to first formalize the abstract ideas behind the algorithm,
    and then refine it to an efficient implementation close to what we had written in C++.
    At this point, one may ask why not directly verify the C++ implementation: while this might be
    possible, the required work and steps would be similar: to manage the complexity of such a verification,
    several bottom-up refinement steps would be necessary, ultimately arriving at something similarly abstract
    as our initial abstract algorithm.

@@

    \subsection{Related Work}
    While there is extensive work on parallel sorting algorithm (e.g.~\cite{AWFS22,CNLM08,AMI21}),
    there seems to be almost no work on their formal verification. The only work we are aware of is
    a distributed merge sort algorithm~\cite{HKBK20}, for which ''no effort has been made to make it efficient''\cite[Sec.~2]{HKBK20},
    nor any executable code has been generated or benchmarked. Another verification~\cite{SaHu20} uses
    the VerCors deductive verifier to prove the permutation property (\is{mset xs' = mset xs})
    of odd-even transposition sort~\cite{Ha72}, but neither the sortedness property nor termination.

>>>>

On top of the back-end, we formalize a concurrent separation logic~\cite{OH04} and implement a verification condition generator (VCG). (Section~\ref{sec:seplogic}).

@@

Using our approach, complex algorithms and data structures can be developed and
refined to optimized efficient code. The stepwise refinement ensures a separation of concerns
between high-level algorithmic ideas and low-level optimizations. We have already used this approach to
verify a wide range of practically efficient sequential algorithms~\cite{ELNN13,BrLa18,WiLa18,La17_CADE,La17_SAT,FBL18,La14,LaSe16,LaSe19,La20}.
In Section~\ref{sec:parsort}, we use our extended techniques to verify a parallel sorting algorithm,
with competitive performance wrt.\ unverified state-of-the-art algorithms.

@@

\subsection{Notation}
Formal statements in this paper correspond to theorems proved in our Isabelle/HOL formalization,
though sometimes simplified to improve the clarity of the presentation.

We mainly use Isabelle/HOL notation, with some shortcuts and adaptations for presentation in a paper.
In this section, we give examples of the more unusual notations: for implication, we always write \is{==>},
e.g., \is{\<forall>x. x<4 ==> x<=4} (in Isabelle, one must use \is{-->} here). Free variables are
universally quantified at the top-level.

Type variables are \is{'a, 'b, ...} and function types are written curried as \is{'a => 'b => 'c}.
Types can be annotated to any term, e.g., \is{(x::nat) < 4 ==> x<=3}.
Function application is written as \is{f x y}, and function update is \is{f(x:=y)}.
For definitions, we use \is{==}, e.g., \is{f x == x+1}.

Algebraic datatypes are written as \is{'a option == Some (the: 'a) | None}. This also defines the selector function \is{the :: 'a option => 'a}.
Values of the tuple type \is{'a_1 \x ... \x 'a$_n$} are written as \is{(a_1,...,a$_n$)}. Lists have type \is{'a list},
and we write \is{[]} for the empty list, \is{x#xs} for the list with head \is{x} and tail \is{xs}, and \is{xs_1@xs_2} for list concatenation.
We also use the list notation \is{[1,2,3]}. The length of list \is{xs} is \is{|xs|}.

The empty set is \is${} :: 'a set$. The set \is${l..<h}$ contains all elements between $l$ inclusive and $h$ exclusive, and the set \is${l..}$ contains all elements greater than or equal to $l$.
Disjoint union is written as \is{$\dot\cup$}. We only write \is{$\dot\cup$} if it is clear from the context where the disjointness constraint belongs to.
Otherwise, we will explicitly write, e.g., \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of the finite set \is{s} is \is{|s|}.

Lambda abstraction is written as \is{\<lambda>x. x+1}. When clear from the context, we omit the \is{\<lambda>}, e.g. \is{spec x. x>42}.
We also use Haskell-like sections for infix operators, e.g. \is{(+1)} for \is{\<lambda>x. x+1}.
We also use underscores to indicate the parameter positions, e.g., \is${_..<_}$ for \is$\<lambda>l h. {l..<h}$.

@@

  Now assume that we have a state (memory) type \is{\mu}, and an access report type \is{\rho}, which forms a monoid (\is{0,+}).
  With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
  \begin{lstlisting}
    'x M == \mu => ('x \x \rho \x \mu) neM
    return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
    bind$_M$ m f \mu == (x_1,r_1,\mu) <- m \mu; (x_2,r_2,\mu) <- f x_1 \mu; return$_{ne}$ (x_2,r_1+r_2,\mu)
  \end{lstlisting}
  Here, \is{return} does not change the state, and reports no accesses (\is$0$),
  and \is{bind} sequentially composes the executions, threading through the state $\mu$ and adding up the access reports \is$r_1$ and \is$r_2$.

  Typically, the access report will contain read and written addresses, such that data races can be detected.
  Moreover, if parallel executions can allocate memory, we must detect those
  executions where the memory manager allocated the same block in both parallel strands.
  As we assume a thread safe memory manager, those \emph{infeasible} executions can safely be ignored.
  Let \is{feasible :: \rho => \rho => bool} and \is{norace :: \rho => \rho => bool} be symmetric predicates,
  and let \is{combine :: (\rho \x \mu) => (\rho \x \mu) => (\rho \x \mu) } be a commutative operator to
  combine two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
  \begin{lstlisting}
    (m_1 || m_2) \mu ==
      (x_1,r_1,\mu_1) <- m_1 \mu; (x_2,r_2,\mu_2) <- m_2 \mu;  --"\rightcomment{execute both strands}"
      assume feasible \rho_1 \rho_2;  --"\rightcomment{ignore infeasible combinations}"
      assert norace \rho_1 \rho_2; --"\rightcomment{fail on data race}"
      return$_{ne}$ ((x_1,x_2), combine (\rho_1,\mu_1) (\rho_2,\mu_2))  --"\rightcomment{combine results}"

    assume P == if P then return () else spec (%_. False)
    assert P == if P then return () else fail
  \end{lstlisting}
  Here, we use \is{assume} to ignore infeasible executions, and \is{assert} to fail on data races.
  Note that, if one parallel strand fails, and the other parallel strand has no possible
  results (\is{spec (\<lambda>_. False)}), the behaviour of the parallel composition is not clear.
  For this reason, we fix an invariant \is{invar$_M$ :: (\mu => ('x \x \rho \x \mu) neM) => bool},
  which implies that every non-failing execution has at least one possible result.
  We define the actual type \is{M} as the subtype satisfying \is{invar$_M$}.
  Thus, we have to prove that every combinator and instruction of our semantics
  preserves the invariant, which is an important sanity check.
  As additional sanity check, we prove symmetry of parallel composition:
  \begin{lstlisting}
    m_1 || m_2 = mswap (m_2 || m_1)     $\text{where}$    mswap m == (x_1,x_2)<-m; return (x_2,x_1)
  \end{lstlisting}
%   where \is|mswap m = (x_1,x_2)<-m; return (x_2,x_1)| swaps the result tuple.

@@

    We then show that \is{malloc} satisfies the invariant of M: we correctly report the allocated block.
    Moreover, we can select any fresh block. As our memory model guarantees an infinite supply of fresh blocks,
    any finite set of blocks can be avoided.
    \begin{lstlisting}
      invarM_malloc: invarM (malloc vs)
    \end{lstlisting}

@@

    Indexing and updating of arrays is implemented by:
    \begin{lstlisting}
      aget p i == ll_ofs_ptr p i; ll_load p
      aset p i x == ll_ofs_ptr p i; ll_store x p; return p
    \end{lstlisting}
    The abstract and concrete programs are linked by the following refinement theorems:
    \begin{lstlisting}
      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ (aget xs$_\dagger$ i$_\dagger$)$\hspace*{-1pt}$ (arr$_A$ xs$_\dagger$ xs$\hspace*{1pt}$*$\hspace*{1pt}$idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ id$_A$$\hspace*{-1pt}$ $\top$$\hspace*{-1pt}$ (lget xs i)
      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i) (aset xs$_\dagger$ i$_\dagger$ x) (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r=xs$_\dagger$) (lset xs i x)
    \end{lstlisting}
    That is, if the list \is$xs$ is refined by array \is{xs$_\dagger$}, and the natural number \is$i$ is refined
    by the fixed-width\footnote{We use Isabelle's word library here, which encodes the actual width as a type variable,
    such that our functions work with any bit width. For code generation, we will fix the width to 64 bit.} word \is{i$_\dagger$} (\is{idx$_A$ i$_\dagger$ i}),
    the \is{aget} operation will return the same result as the \is{lget} operation (\is{id$_A$}).
    The resulting memory will still contain the original array. Note that there is no explicit precondition that the array access
    is in bounds, as this follows already from the assertion in the abstract \is{lget} operation.
    The \is{aset} operation will return a pointer to an array that refines the updated list returned by \is{lset}.
    As the array is updated in place, the original refinement of the array is no longer valid.
    Moreover, the returned pointer \is{r} will be the same as the argument pointer \is{xs$_\dagger$}.
    This information is important for refining to parallel programs on disjoint parts of an
    array (cf.\ Section~\ref{sec:array_split}).

@@

  Next, we implement a set as the union of a list of non-empty, pairwise disjoint, and finite sets.
  While that seems to make little sense at first glance, we will later implement the sets in the list by intervals,
  and the list itself by dynamic arrays, to obtain an imperative interval list data structure.
  We define operations for constructing an empty set, emptiness test,
  disjoint union with a single set, cardinality, and a more specialized operation \is{split}, which splits off
  a non-empty set from the list:
  \begin{lstlisting}
    ivl$_R$ :: nat set list => nat set => bool
    ivl$_R$ ls s == s = \<Union>set s  &  pw_disjoint ls  &  (\<forall>x:set ls. finite x & x ~= {})

@@

  We then refine the lists of sets to array lists (dynamic arrays) of intervals: \is{al$_A$ iv$_A$}.
  Here, \is{al$_A$} is the refinement assertion from lists to the array list data structure from the IRF collections library.
  As argument, it takes the refinement relation for the list elements.
  Again, Sepref automatically generates imperative implementations of the \is{ivl} functions and proves the corresponding
  refinement lemmas. Combining them with the refinements to sets yields the desired imperative interval list data structure,
  with the refinement relation \is{ivl$_A$ == al$_A$ iv$_A$ O ivl$_R$}.
  For example, for joining a single interval to the list, and for splitting off an interval, we get:
  \begin{lstlisting}
    ivl_dj_un$_\dagger$, ($\dot\cup$) : iv$_A$ -> ivl$_A$^d -> ivl$_A$
    ivl_split$_\dagger$, split : ivl$_A$^d -> iv$_A$ \x ivl$_A$
  \end{lstlisting}
  Note that we update the underlying dynamic array destructively, hence the $\cdot^d$ annotation to the argument refinements.

@@

  Analogously to \is{with_split}, we define a combinator \is{with_idxs s xs f}, that splits
  the list \is{xs} into the lists with the indexes \is{s}, and without the indexes \is{s}, executes \is{f} on these lists,
  and joins the resulting lists:
  \begin{lstlisting}
  with_idxs s xs f ==
    assert (\<forall>i\<in>s. i<|xs| & xs!i ~= None);
    let (xs_1, xs_2) = (sl_split s xs, sl_split (-s) xs);
    (xs$_1'$,xs$_2'$) <- f xs_1 xs_2;
    assert (sl_struct xs$_1'$ = sl_struct xs_1 & sl_struct xs$_2'$ = sl_struct xs_2);
    return (sl_join xs$_1'$ xs$_2'$)
  \end{lstlisting}

@@

  Figure~\ref{fig:psort} shows our abstract parallel sorting algorithm \is{psort}.
  This algorithm is derived from the well-known quicksort and introsort algorithms~\cite{Muss97}:
  like quicksort, it partitions the list (line~\ref{l:psort:partition}), and then recursively sorts the partitions in parallel (l.~\ref{l:psort:rec_par}).
  Like introsort, when the recursion gets too deep, or the list too short, we fall back to
  some (not yet specified) sequential sorting algorithm (l.~\ref{l:psort:fallback}). Similarly, when the partitioning is very unbalanced (l.~\ref{l:psort:bad}),
  we sort the partitions sequentially (l.~\ref{l:psort:rec_seq}). These optimizations aim at not spawning threads for small sorting tasks, where
  the overhead of thread creation outweighs the advantages of parallel execution.
  A more technical aspect is the extra parameter \is{n} that we introduced for the list length.
  Thus, we can refine the list to just a pointer to an array, and still access its length\footnote{Alternatively, we could refine a list to a pair of array pointer and length.}.

@@

  Our sorting algorithms are developed and verified in the context of this locale (to avoid confusion, our presentation has, up to now, just used \is{<}, \is{<=}, and \is{sorted} instead of \is{lt p}, \is{le p}, and \is{sorted_wrt (le p)}).
  To get an actual sorting algorithm, we instantiate the locale with an abstract and concrete compare function, proving
  that the abstract function is a weak ordering, and that the concrete function refines the abstract one.
  For our example of sorting indexes into an array, where the array elements themselves are compared by a function \is{lt}, we get:
  \begin{lstlisting}
    lt_idx xs i j == lt (xs!i) (xs!j)
    lt_idx$_\dagger$ xs$_\dagger$ i$_\dagger$ j$_\dagger$ == x$_\dagger$<-aget xs$_\dagger$ i$_\dagger$; y$_\dagger$<-aget xs$_\dagger$ j$_\dagger$; lt$_\dagger$ x$_\dagger$ y$_\dagger$

    interpretation idx: pcmp lt_idx lt_idx$_\dagger$ arr$_A$ idx$_A$   \<proof>
  \end{lstlisting}
  This yields sorting algorithms for sorting indexes, taking an extra parameter for the array to index into.
  For our sampling application, we use \is{idx.introsort xs}.

@@

  \begin{tikzpicture}
    \begin{loglogaxis}[
      xlabel near ticks,
      legend style = {
        at = {(.97,.4)},
        cells={anchor=west},
        font=\footnotesize
      },
      ymajorgrids,
      xmin=10^4,xmax=10^8,
      title={\large Server (u64)},
      title style={at={(0.25,.75)}},
      width=.54\textwidth,
      height=.3\textheight
    ]
      \input{sadata_uint64_server}
    \end{loglogaxis}
  \end{tikzpicture}\hfill
  \begin{tikzpicture}
    \begin{loglogaxis}[
      xlabel near ticks,
      legend style = {
        at = {(.97,.4)},
        cells={anchor=west},
        font=\footnotesize
      },
      ymajorgrids,
      xmin=10^3,xmax=10^7,
      title={\large Server (str)},
      title style={at={(0.25,.75)}},
      width=.54\textwidth,
      height=.3\textheight
    ]
      \input{sadata_llstring_server}
      \legend{}
    \end{loglogaxis}
  \end{tikzpicture}
  \caption{Runtimes for sorting small arrays of randomly distributed integers and strings.
    The $y$-axis shows the runtime in milliseconds, and the $x$-axis the array size. Note that both axis are logarithmic.
  }\label{fig:smallarrays}

@@

    While the verification of our sorting algorithms uses a top-down approach, we actually started with
    implementing, benchmarking, and fine-tuning the algorithms in C++. This gave us a quick way to
    find an algorithm that is efficient, and, at the same time not too complex for verification,
    without having to verify each intermediate step towards this algorithm.
    Only then we used our top-down approach to first formalize the abstract ideas behind the algorithm,
    and then refine it to an efficient implementation close to what we had written in C++.
    At this point, one may ask why not directly verify the C++ implementation: while this might be
    possible, the required work and steps would be similar: to manage the complexity of such a verification,
    several bottom-up refinement steps would be necessary, ultimately arriving at something similarly abstract
    as our initial abstract algorithm.

@@

    \subsection{Related Work}
    While there is extensive work on parallel sorting algorithms (e.g.~\cite{AWFS22,CNLM08,AMI21}),
    there seems to be almost no work on their formal verification. The only work we are aware of is
    a distributed merge sort algorithm~\cite{HKBK20}, for which ''no effort has been made to make it efficient''\cite[Sec.~2]{HKBK20},
    nor any executable code has been generated or benchmarked. Another verification~\cite{SaHu20} uses
    the VerCors deductive verifier to prove the permutation property (\is{mset xs' = mset xs})
    of odd-even transposition sort~\cite{Ha72}, but neither the sortedness property nor termination.
